---
title: "NFL Web Scraping"
author: Igor Aboim 
output:
  md_document:
    variant: markdown_github
---
### Web Scraping 2019 NFL game logs

In this project there is an example of web scraping using R (robotstxt and rvest).

Loading Packages

```{r}
library(robotstxt)
library(rvest)
library(stringr)
library(plyr)
library(tidyverse)
```

Confirming using robotstxt that it is acceptable to scrape from the website.

```{r}
base_url = "https://www.pro-football-reference.com"
paths_allowed(base_url)
```

Writing a script that will output the links to game log data for all 30 teams. As an example, the link for the Dallas Cowboys is https://www.pro-football-reference.com/teams/dal/2019.htm.

```{r}
####Modify file paths to get game logs
pages <- read_html(file.path(base_url)) %>%
  html_nodes("#teams .left a") %>%
  html_attr("href") %>%
  file.path(base_url, .) %>%
  str_replace("2020", "2019") %>%
  str_replace(".htm", "/gamelog/")

pages
```

Writing a script that will grab the game log data for a specific team and its respective opponent.

```{r}
gamelog.temp <- read_html(pages[1]) %>%
  html_nodes("#gamelog2019") %>%
  html_table(header=T)
gamelog.temp <- gamelog.temp[[1]]
opp.gamelog.temp <- read_html(pages[1]) %>%
  html_nodes("#gamelog_opp2019") %>%
  html_table(header=T)
opp.gamelog.temp <- opp.gamelog.temp[[1]]
```

Cleaning up the data - remove redundant rows, ensure column names don’t include special characters, ensure all numeric columns are actually numeric.

```{r, warning=FALSE}
colnames(gamelog.temp) <- c("week", "day", "date", "boxscore_word", "game_outcome", "overtime", "game_location", "opp", "points_scored", "points_allowed", "pass.cmp", "pass.att", "pass.yds", "pass.TD", "pass.int", "pass.sk", "pass.yds", "pass.Y/A", "pass.NY/A", "pass.cmp%", "pass.rate", "rush.att", "rush.yds", "rush.Y/A", "rush.TD", "FGM", "FGA","XPM", "XPA", "punt.pnt", "punt.yds", "3Dconv", "3DAtt", "4DConv", "4DAtt", "ToP")
colnames(opp.gamelog.temp) <- c("week", "day", "date", "boxscore_word", "game_outcome", "overtime", "game_location", "opp", "points_scored", "points_allowed", "opp.pass.cmp", "opp.pass.att", "opp.pass.yds", "opp.pass.TD", "opp.pass.int", "opp.pass.sk", "opp.pass.yds", "opp.pass.Y/A", "opp.pass.NY/A", "opp.pass.cmp%", "opp.pass.rate", "opp.rush.att", "opp.rush.yds", "opp.rush.Y/A", "opp.rush.TD", "opp.FGM", "opp.FGA","opp.XPM", "opp.XPA", "opp.punt.pnt", "opp.punt.yds", "opp.3Dconv", "opp.3DAtt", "opp.4DConv", "opp.4DAtt", "opp.ToP")

gamelog.temp <- gamelog.temp[-1,]
gamelog.temp[,c(11:25)] <- sapply(gamelog.temp[,c(9:35)], as.numeric)
gamelog.temp$team <- str_sub(pages[1], 47, 49)

opp.gamelog.temp <- opp.gamelog.temp[-1,]
opp.gamelog.temp[,c(11:25)] <- sapply(opp.gamelog.temp[,c(9:35)], as.numeric)
opp.gamelog.temp$team <- str_sub(pages[1], 47, 49)
```

Concatenating Gamelog and Opposition Gamelog dataframes

```{r, warning=FALSE, message=FALSE}
nwe_gamelog <- join(gamelog.temp, opp.gamelog.temp,type="full")
head(nwe_gamelog)
```

Systematically collecting all game logs for all the teams.

```{r, warning=FALSE, message=FALSE}
####First, create a placeholder for each team's game logs
gamelog.team <- list()

####Function
for(i in 1:length(pages)){
  
  gamelog.temp <- read_html(pages[i]) %>%
    html_nodes("#gamelog2019") %>%
    html_table(header=T)
  gamelog.temp <- gamelog.temp[[1]]
  opp.gamelog.temp <- read_html(pages[i]) %>%
    html_nodes("#gamelog_opp2019") %>%
    html_table(header=T)
  opp.gamelog.temp <- opp.gamelog.temp[[1]]
  
  colnames(gamelog.temp) <- c("week", "day", "date", "boxscore_word", "game_outcome", "overtime", "game_location", "opp", "points_scored", "points_allowed", "pass.cmp", "pass.att", "pass.yds", "pass.TD", "pass.int", "pass.sk", "pass.yds", "pass.Y/A", "pass.NY/A", "pass.cmp%", "pass.rate", "rush.att", "rush.yds", "rush.Y/A", "rush.TD", "FGM", "FGA","XPM", "XPA", "punt.pnt", "punt.yds", "3Dconv", "3DAtt", "4DConv", "4DAtt", "ToP")
  colnames(opp.gamelog.temp) <- c("week", "day", "date", "boxscore_word", "game_outcome", "overtime", "game_location", "opp", "points_scored", "points_allowed", "opp.pass.cmp", "opp.pass.att", "opp.pass.yds", "opp.pass.TD", "opp.pass.int", "opp.pass.sk", "opp.pass.yds", "opp.pass.Y/A", "opp.pass.NY/A", "opp.pass.cmp%", "opp.pass.rate", "opp.rush.att", "opp.rush.yds", "opp.rush.Y/A", "opp.rush.TD", "opp.FGM", "opp.FGA","opp.XPM", "opp.XPA", "opp.punt.pnt", "opp.punt.yds", "opp.3Dconv", "opp.3DAtt", "opp.4DConv", "opp.4DAtt", "opp.ToP")
  
  gamelog.temp <- gamelog.temp[-1,]
  gamelog.temp[,c(11:25)] <- sapply(gamelog.temp[,c(9:35)], as.numeric)
  gamelog.temp$team <- str_sub(pages[i], 47, 49)
  
  opp.gamelog.temp <- opp.gamelog.temp[-1,]
  opp.gamelog.temp[,c(11:25)] <- sapply(opp.gamelog.temp[,c(9:35)], as.numeric)
  opp.gamelog.temp$team <- str_sub(pages[i], 47, 49)
  
  gamelog <- join(gamelog.temp, opp.gamelog.temp,type="full")
  gamelog.team[[i]] <- gamelog
}

```

Compilation of all game logs into a single data frame.

```{r}
gamelogs.all <- rbind.fill(gamelog.team)
```

Creating the final data set “nfl2019_gamelogs”. Output of the results for str(nfl2019_gamelogs)

```{r}
nfl2019_gamelogs <- gamelogs.all
str(nfl2019_gamelogs)
```
